---
title: "An Introduction to tseffects: Robust Dynamic (Causal) Inferences from Single-Equation Time Series Models (with Interactions)"
author: "Soren Jordan, Garrett N. Vande Kamp, and Reshi Rajan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: timeseries.bib
vignette: >
  %\VignetteIndexEntry{An Introduction to tseffects: Robust Dynamic (Causal) Inferences from Single-Equation Time Series Models (with Interactions)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Autoregressive distributed lag (A[R]DL) models (and their reparameterized equivalent, the Generalized Error-Correction Model [GECM]) are the workhorse models in uncovering dynamic inferences. Since the introduction of the general-to-specific modeling strategy of Hendry [-@hendry], users have become accustomed to specifying a single-equation model with lags of the dependent variable included alongside the independent variables (in differences or levels, as appropriate) to capture dynamic inferences. Through all of the follolwing, we assume users are familiar with such models and their best practices, including testing for stationarity on both sides of the equation [@philipsavoid], equation balance [@pickupkellstedt19], the absence of serial autocorrelation in the residuals, or the testing for moving averages [@gvkjordan]. 

ADL models are simple to estimate; this is what makes them attractive. Once these models are estimated, what is less clear is how to uncover a rich set of dynamic *inferences* from these models. This complication arises in two key areas. First: despite robust work on possible quantities of interest that can be derived from dynamic models (see De Boef and Keele [-@de2008taking]), the actual calculation of these quantities (and corresponding measures of uncertainty) remains a considerable difficulty for practitioners. Formulae are often complex and model-dependent (i.e. on the underlying lag structure of the model). While other, simulation-based methods exist to interpret A[R]DL models (as in Jordan and Philips [-@JordanPhilips2018RJ,-@jordan2018cointegration]), these methods are not well equipped to test hypotheses of interest.

Second: these inferences are rarely causal. While some have developed frameworks for causal inference using alternative dynamic models [@Blackwell2013], little work recognizes the potential of the A[R]DL model itself to provide causal inferences. Vande Kamp, Jordan, and Rajan [-@gvkjr] begin to fill this gap by specifying the conditions under which the ADL model provides causal inferences, but these inferences too are difficult to calculate. 

`tseffects` solves both problems. To show how, we first re-introduce the ADL and GECM models and define typical effects calcualted from these models. We then apply those effects to each of the unique problems outlined above.

# The ADL/GECM model

To help ensure consistency in notation, we define the ADL($p$,$q$) model as follows

$$
    Y_{t} = \alpha_{0} + \sum_{i = 1}^{p}\alpha_{i}Y_{t-i} + \sum_{j = 0}^{q}\beta_{j}X_{t-j} + \sum_{j = 0}^{q}\gamma_{j}Z_{t-j} + \sum_{k = 0}^{r}\theta_{k}\varepsilon_{t-k}
$$

Similarly, we define its GECM equivalent as

$$
    \Delta Y_{t} = \alpha_{0} + \sum_{i = 1}^{p-1}\alpha_{i}\Delta Y_{t-i} + \theta_{0}Y_{t-1} + \sum_{j = 0}^{q-1}\beta_{j}\Delta X_{t-j} + \theta_{1}X_{t-1} + \sum_{j = 0}^{q}\gamma_{j}Z_{t-j} + \sum_{k = 0}^{r}\theta_{k}\varepsilon_{t-k}
$$

Notice in particular the form of the GECM with a single first lag of both the dependent and independent variable in levels. (The GECM calculations provided by `ts.ci.gecm.plot` assume this correctly specified GECM and use the variable translations between the GECM and ADL to define the coefficient in ADL terms.) We will use these models to define and recover both causal effects and traditional time-series effects from the underlying model. 



# Causal inferences with `ts.effects`

We define the causal estimand of the Pulse Treatment Effect (PTE), commonly referred to as an impulse response function (or perhaps a short-run effect) in the time series literature. 

$$PTE_{h} = \beta_{h} + \sum_{i=1}^{h}\alpha_{i}PTE_{h-i}$$

We additionally define the causal estimand of the Step Treatment Effect (STE), commonly referred to as a cumulative impulse response function in the time series literature.

$$STE_{h} = \sum_{k=0}^{h}PTE_{k} = PTE_{h} + STE_{h-1}$$

These quantities are often estimated from an ADL including a dependent variable in levels or differences, an independent variable in levels or differences, or a mix of both. Yet scholars might want to make causal inferences regarding the dependent variable in levels *regardless* of its order of differencing in the ADL model. Or, scholars might want to uncover the effects of a treatment applied to an independent variable in levels, even if that independent variable is entered into the ADL in differences. Regular ADL formulae cannot deliver on this expectation. To resolve this deficiency, Vande Kamp, Jordan, and Rajan [-@gvkjr] introduce the General Dynamic Treatment Effect

$$GDTE_{h,n,d.y,d.x} = E[\Delta^{d.y}Y_{t+h}|\Delta^{d.x}X_{t:t+h} = \Delta^{d.x}x_{t:t+h} + c_{0:h,n}] - E[\Delta^{d.y}Y_{t+h}|\Delta^{d.x}X_{t:t+h}= \Delta^{d.x}x_{t:t+h}]$$
with the following elements: 

- $d.y$: the number of times the independent variable is differenced *before* determining the effect of the treatment on the dependent variable
- $d.x$: the number of times the independent variable is differenced *before* a treatment is applied to it
- $h$: the number of time periods since a treatment was applied

$n$ deserves special explanation. [@gvkjr] demonstrate how the binomial function can be used to represent traditional and beyond-traditional treatment histories typically examined in time series

- $n = 1$: Treatment sequence $(1,2,3,4,\dots)$; Linear Trend Coefficient of $X$
- $n = 0$: Treatment sequence $(1,1,1,1,\dots)$; Permanent One-Unit Increase (Step)
- $n = -1$: Treatment sequence $(1,0,0,0,\dots)$ One-Time, One-Unit Increase (Pulse)

The leverage of the GDTE is that it "translates" these popular treatment histories to recover inferences in the desired form of the independent and dependent variables, regardless of their level of differencing, given the specified treatment history. All scholars need to do is estimate the ADL model and choose a treatment.

## Two subfunctions: `pte.calculator` and `GDTE.calculator`

While its unlikely that users will want to interact with these "under the hood" functions, they might be useful either for teaching or for developing a better understanding of the GDTE. Despite their names, neither function actually does any "calculation"; rather, they return an `mpoly` class of the formula that would be evaluated.

The first is `pte.calculator`. Like its name implies, this calculates the traditional PTE for $h = 0$ up to $h = $ some `limit`. It returns a list of formulae, each one the PTE for the relevant period. (Time series scholars might refer to these as the period-specific effects.) It constructs these formulae given a named vector of the relevant $x$ variables (passed to `x.vrbl`) and their corresponding lag orders, as well as the same for $y$ (passed to `y.vrbl`). For instance, if you estimate an ADL(1,2), where the $x$ variables are `mood`, `l1.mood`, and `l2.mood` (lags need not be consecutive, though in a well-specified ADL they likely are) and the dependent variable is `policy` (with its first lag `l1.policy`), `pte.calculator` would look like

```{r}
library(tseffects)
ADL1.2.ptes <- pte.calculator(x.vrbl = c("mood" = 0, "l1_mood" = 1, "l2_mood" = 2), 
               y.vrbl = c("l1_policy" = 1),
               limit = 5)
```

This list is six elements long: including the PTE for each period from 0 to 5. These PTEs form the basis of the GDTE, given the order of differencing of $x$ and $y$ and the specified treatment history $n$. To create the formulae for the GDTE, `GDTE.calculator` requires five arguments, most of which are defined above: `d.x` and `d.y` (defined with respect to the GDTE), `n` (the treatment history applied), `limit`, and `pte`. The novel argument, `pte`, is the list of PTEs for each period in which a GDTE is desired. 

Assume the same ADL(1,2) from above. To recover the formulae for a GDTE of a pulse treatment, assuming that both `mood` and `policy` are in levels (such that both `d.x` and `d.y` are 0), we would run

```{r}
GDTE.calculator(d.x = 0, d.y = 0, n = -1, limit = 5, pte = ADL1.2.ptes)
```

(Since `GDTE.calculator` is "under the hood," `n` can only be specified as an integer. Later, in a more user-friendly function, we will see that treatment histories can also be specified using characters.) 

Recall that we might want to generalize our inferences beyond simple ADLs with both variables in levels. For exposition, assume that `mood` was integrated, so, to balance the ADL, it needed to be entered in differences. The user does this and enters $x$ variables as `d.mood`, `l1.d.mood`, and `l2.d.mood`. `pte.calculator` would look like

```{r}
ADL1.2.d.ptes <- pte.calculator(x.vrbl = c("d_mood" = 0, "l1_d_mood" = 1, "l2_d_mood" = 2), 
               y.vrbl = c("l1_policy" = 1),
               limit = 5)
```

`tseffects` allows the user to recover inferences of a treatment history applied to $x$ in levels, rather than its observed form of differences. All we need to do is specify the `d.x` order.


```{r}
GDTE.calculator(d.x = 1, d.y = 0, n = -1, limit = 5, pte = ADL1.2.d.ptes)
```

As [@gvkjr] suggest, scholars should be very thoughtful about the treatment history applied. These treatments are applied to the underlying independent variable *in levels*. Since $x$ is integrated in this example, it might be more reasonable to imagine that the underlying levels variable is actually trending. To revise this expectation and apply a new treatment history, we would run 

```{r}
GDTE.calculator(d.x = 1, d.y = 0, n = 1, limit = 5, pte = ADL1.2.d.ptes)
```

(Notice that the PTEs do not change.) Recall that these are "under the hood": we expect instead that most users will interact with the full calculation and plotting functions, discussed below.


## User-friendly causal effects: `ts.ci.adl.plot` and `ts.ci.gecm.plot`

Again, we don't expect most users will not interact with these calculators. Instead, they will estimate an ADL model via `lm` and expect causal effects. To achieve this directly, we introduce `ts.ci.adl.plot`. We gloss over arguments that we've already seen in `pte.calculator` and `GDTE.calculator`:

- `x.vrbl`: a named vector of the $x$ variable and its lag orders
- `y.vrbl`: a named vector of the $y$ variable and its lag orders
- `d.x`: the order of differencing of the independent variable before a treatment is applied
- `d.y`: the order of differencing of the dependent variable before causal effects are recovered
- `h.limit`: the number of periods for which to calculate the treatment effects (the default is 20)

Now, since effects will be calcualted, users must also provide a `model`: the `lm` object containing the (well-specified, etc. ADL model) that has the estimates. Users must also specify a treatment history through `te.type`. Users can continue to specify an integer that represents `n` or can use the phrases `pte` or `pulse` (for $n = -1$) and `ste` or `step` (for $n = 0$). To generalize the challenge of variables entered in differences, users can specify `inferences.y` and `inferences.x`. If both $x$ and $y$ are entered into the ADL in levels (so that `d.x` and `d.y` are 0), `inferences.y` and `inferences.x` default to `levels`. If, however, either $x$ or $y$ (or both) are entered in differences, users can recover causal effects from a treatment history applied to $x$ in levels, or to a dependent variable realized in levels. However, *they don't have to*. The treatment can still be applied to $x$ in differences, or the causal effects can be realized on $y$ in differences: just change `inferences.x` to `differences` (for the former) or `inferences.y` to `differences` (for the latter).

A few options allows users to tailor levels and types of uncertainty desired. The first is the type of standard errors calculated from the model: users can specify any type of standard error accepted by `vcovHC` in the `sandwich` package: this is done by `se.type = ` (where `const` is the default). Users can also specify a level of uncertainty to the delta method other than 0.95; this is done through `dM.level`.

Three final arguments control the output returned to the user. 

  - `return.plot`: we assume most users would like a visualization of these results and nothing else (default is `TRUE`)
  - `return.data`: if the user wants the raw data (to design their own `ggplot`, for instance), this argument should be `TRUE` (default is `FALSE`)
  - `return.formulae`: the formulae can be very complex, but perhaps users want to see the particular formula for the effect in some particular period $h$. If `return.formulae = TRUE`, these are returned as a list, along with the binomial series that represents the treatment history (default is `FALSE`)

We'll use the toy time series data included with the package to illustrate the function. These data are purely expository, as is the model. We do not advocate for this model. 

```{r}
data(toy.ts.interaction.data)

# Fit an ADL(1, 1)
model.adl <- lm(y ~ l_1_y + x + l_1_x, data = toy.ts.interaction.data)

test.pulse <- ts.ci.adl.plot(model = model.adl,
                                   x.vrbl = c("x" = 0, "l_1_x" = 1), 
                                   y.vrbl = c("l_1_y" = 1),
                                   d.x = 0, 
                                   d.y = 0,
                                   te.type = "pulse", 
                                   inferences.y = "levels", 
                                   inferences.x = "levels",
                                   h.limit = 20, 
                                   return.plot = TRUE, 
                                   return.formulae = TRUE)
  
```

This recovers the pulse treatment effect. We can also observe the formulae and associated binomials in `test.pulse$formulae` and `test.pulse$binomials`. `test.pulse$plot` contains the plot: the effect of applying a pulse treatment to $x$. If we wanted the STE, we would change `te.type`.

```{r}
test.pulse2 <- ts.ci.adl.plot(model = model.adl,
                                   x.vrbl = c("x" = 0, "l_1_x" = 1), 
                                   y.vrbl = c("l_1_y" = 1),
                                   d.x = 0, 
                                   d.y = 0,
                                   te.type = "ste", 
                                   inferences.y = "levels", 
                                   inferences.x = "levels",
                                   h.limit = 20, 
                                   return.plot = TRUE, 
                                   return.formulae = TRUE)
  
```

Equivalently, we could specify `te.type` = `step` or `0` (since this is the value of $n$ that represents a step treatment history).

Imagine that the $x$ variable was actually integrated, so, for balance, it was entered into the model in differences. 

```{r}
data(toy.ts.interaction.data)

# Fit an ADL(1, 1)
model.adl.diffs <- lm(y ~ l_1_y + d_x + l_1_d_x, data = toy.ts.interaction.data)
```

To observe the causal effects of a pulse treatment history applied to $x$ *in levels*, we would run

```{r}
ts.ci.adl.plot(model = model.adl.diffs,
                                   x.vrbl = c("d_x" = 0, "l_1_d_x" = 1), 
                                   y.vrbl = c("l_1_y" = 1),
                                   d.x = 1, 
                                   d.y = 0,
                                   te.type = "pte", 
                                   inferences.y = "levels", 
                                   inferences.x = "levels",
                                   h.limit = 20, 
                                   return.plot = TRUE)
```

(Notice the change in `d.x` and the `x.vrbl`). Suppose, though, that we wanted to apply this pulse treatment to $x$ *in differences*. We just change the form of `inferences.x`, leaving all other arguments the same.

```{r}
ts.ci.adl.plot(model = model.adl.diffs,
                                   x.vrbl = c("d_x" = 0, "l_1_d_x" = 1), 
                                   y.vrbl = c("l_1_y" = 1),
                                   d.x = 1, 
                                   d.y = 0,
                                   te.type = "pte", 
                                   inferences.y = "levels", 
                                   inferences.x = "differences",
                                   h.limit = 20, 
                                   return.plot = TRUE)
```

The same changes could be made to `inferences.y` (if it  were modeled in differences). Generally speaking: `d.x` should be entered as the form of the independent variable, even if the user desires treatment effects recovered in the differenced form (this should be handled by `inferences.x`). The same is true of `d.y`.

`ts.ci.gecm.plot` is extremely similar. Imagine the following specified GECM model

```{r}
# Fit a GECM(1, 1)
model.gecm <- lm(d_y ~ l_1_y + l_1_d_y + l_1_x + d_x + l_1_d_x, data = toy.ts.interaction.data)
```

Since both the $x$ and $y$ variables appear in levels and differences, we need to differentiate them. `ts.ci.gecm.plot` does this by separating them into two forms

- `x.vrbl`: a named vector of the x variables (of the lower level of differencing, usually in levels d = 0) and corresponding lag orders in the GECM model
- `y.vrbl`: a named vector of the (lagged) y variables (of the lower level of differencing, usually in levels d = 0) and corresponding lag orders in the GECM model
- `x.vrbl.d.x`: the order of differencing of the x variable (of the lower level of differencing, usually in levels d = 0) in the GECM model
- `y.vrbl.d.y`: the order of differencing of the y variable (of the lower level of differencing, usually in levels d = 0) in the GECM model
- `x.d.vrbl`: a named vector of the x variables (of the higher level of differencing, usually first differences d = 1) and corresponding lag orders in the GECM model
- `y.d.vrbl`: a named vector of the y variables (of the higher level of differencing, usually first differences d = 1) and corresponding lag orders in the GECM model
- `x.d.vrbl.d.x`: the order of differencing of the x variable (of the higher level of differencing, usually first differences d = 1) in the GECM model
- `y.d.vrbl.d.y`: the order of differencing of the y variable (of the higher level of differencing, usually first differences d = 1) in the GECM model

The orders of differencing of `y.vrbl` and `y.d.vrbl` (as well as `x.vrbl` and `x.d.vrbl`) must be separated by only one order. From here, all of the other arguments are the same. (Unlike in `ts.ci.gecm.plot`, `inferences.x` and `inferences.y` must be in levels, given the dual nature of the variables in the model.) For `model.gecm`, the syntax would be

```{r}
gecm.pulse <- ts.ci.gecm.plot(model = model.gecm,
                                   x.vrbl = c("l_1_x" = 1), 
                                   y.vrbl = c("l_1_y" = 1),
                                   x.vrbl.d.x = 0, 
                                   y.vrbl.d.y = 0,
                                   x.d.vrbl = c("d_x" = 0, "l_1_d_x" = 1),
                                   y.d.vrbl = c("l_1_d_y" = 1),
                                   x.d.vrbl.d.x = 1,
                                   y.d.vrbl.d.y = 1,
                                   te.type = "pulse", 
                                   inferences.y = "levels", 
                                   inferences.x = "levels",
                                   h.limit = 20, 
                                   return.plot = TRUE, 
                                   return.formulae = TRUE)
```

Examining the formulae show that the conversion from GECM coefficients to ADL coefficients is automatic. As before, if we wanted a different treatment history, we would just adjust `te.type` to apply a different treatment history to $x$ in levels.






# Concluding thoughts

A word of caution: `tseffects` is very naive to your model. This is true in three ways. First, it relies on the user to specify the lag structure correctly in both the underlying `lm` model as well as the corresponding plot you are producing. Including a term in the model, but not passing it to the plotting function (i.e. to `ts.ci.adl.plot` through `x.vrbl`) will cause the calculation of the effects to be inaccurate. So, be sure to always include all forms of both $x$ and $y$ constructing effects estimates.

Second, and to reiterate, it makes no check that your model is well specified. The best starting place for a calculation of effects is always a well-specified model; users should complete this before obtaining inferences. See the citations at the beginning of the vignette for places to start.

Third, if using the causal inference functions, it does not pass judgement on the treatment history you are applying. Stationary (`d.x = 0`) variables cannot trend by definition, but specifying `te.type = 1` in the function would apply a linear trend treatment history anyway. This is not a good idea and is inconsistent with the underlying independent variable. However, treatments are counterfactuals, and we are loathe to limit the ability of a user to specify a counterfactual, consistent or not. We leave it in the user's hands to determine if their treatment is sensible or not. 

Despite these words of caution, `tseffects` hopefully simplifies the job of the time series analyst in obtaining a variety of effects.





# Bibliography


